Targeted interrogation of mutational hotspots in rheumatoid arthritis

### 2024-03-06 scomatic/reports/targeted_mutation_calling.Rmd
fix up bam mutation calling

@@

          hotspots %>%
            purrr::pmap(function(chr, pos, mut_pos, gene, ...) {
              deepSNV::bam2R(
                ct_bam,
                chr = chr,
                start = pos,
                stop = pos,
                q = min_phred,
                mask = sam_flag,
                mq = mapq) %>%
                tibble::as_tibble() %>%
                dplyr::mutate(chr = chr, pos = pos, mut_pos = mut_pos, gene = gene) %>%
                dplyr::select(chr, pos, mut_pos, gene, dplyr::all_of(nucleotides))
            }) %>%
            dplyr::bind_rows() %>%
            dplyr::rowwise() %>%
            dplyr::mutate(
              total = sum(dplyr::across(dplyr::all_of(nucleotides)))) %>%
            dplyr::relocate(dplyr::all_of(nucleotides), .after = dplyr::last_col())

@@

```{r plot}
# plot positions
unique(sites$gene) %>%
  purrr::walk(function(i) {
    p <-
      sites %>%
      dplyr::filter(gene == i) %>%
      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                          values_to = 'nt_n_reads', names_to = 'nt') %>%
      dplyr::mutate(
        strandless_nt = 
          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
          forcats::fct_relevel(unique(c(paste0(topupper(nucleotides), '_ref'),
                                        paste0(toupper(nucleotides), '_alt'))))) %>%
      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
      geom_vline(aes(xintercept = mut_pos)) +
      geom_col(position = 'stack') +
      theme_bw() +
      theme(axis.text.x = element_blank()) +
      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free') +
      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
                                   C_ref = 'darkblue', C_alt = 'lightblue',
                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
                                   T_ref = 'darkred', T_alt = 'salmon',
                                   INS_alt = 'black', DEL_alt = 'grey')) +
      ggtitle(i)
    print(p)
    })
```

>>>>

```{r}
# read targeted nanoseq 
tnanoseq <-
  '/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/dnds_nano_muts.tsv' %>%
  readr::read_tsv()

# read genes of interest
target_genes <-
  '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/Sanger_TERT-v4_TE-95148282_hg19_highstringencyfilter_buccal_gene_list.tsv' %>%
  readr::read_tsv()

# run dndscv
dnds <-
  dndscv::dndscv(
    unique(tnanoseq[, 1:5]),
    max_muts_per_gene_per_sample = Inf,
    max_coding_muts_per_sample = Inf,
    gene_list = target_genes$gene,
    outmats = T)

# run sitednds
sitednds <-
  dndscv::sitednds(dnds)

# get the significant genes
signif_genes <-
  dnds$sel_cv %>%
  tibble::as_tibble() %>%
  dplyr::filter(qglobal_cv < 0.1)

# get the sites with positively selected mutations
recursites <-
  sitednds$recursites %>%
  tibble::as_tibble()

# convert to granges (1-based!)
gr <- GRanges(
  seqnames = paste0('chr', recursites$chr),
  ranges = IRanges(start = recursites$pos, end = recursites$pos),
  gene = recursites$gene,
  pos_GRCh37 = recursites$pos)

# get chain file
chain <-
  rtracklayer::import.chain('data/liftOver/hg19ToHg38.over.chain')

# lift over
lifted <-
  rtracklayer::liftOver(gr, chain) %>%
  unlist() %>%
  tibble::as_tibble() %>%
  dplyr::transmute(chr = gsub('chr', '', seqnames), 
                   pos = start, gene, pos_GRCh37)

# update positions
recursites <-
  recursites %>%
  dplyr::rename(pos_GRCh37 = pos) %>%
  dplyr::left_join(lifted, by = c('chr', 'gene', 'pos_GRCh37')) %>%
  dplyr::select(chr, pos, gene, ref, mut, pos_GRCh37, everything())
```

@@

          recursites %>%
            # dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% 
            purrr::pmap(function(chr, pos, start20, stop20, gene, ...) {
              # testing: # recursites %>% dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*', pos == mut_pos) %>% as.list() %>% list2env(envir = globalenv())

              # positions
              positions <- (pos - 20):(pos + 20)

              # get reads at each position of the window
              deepSNV::bam2R(
                ct_bam,
                chr = chr,
                # extend pileup window 20bp either side of mutated position
                start = pos - 20,
                stop = pos + 20,
                q = min_phred,
                mask = sam_flag,
                mq = mapq) %>%
                tibble::as_tibble() %>%
                dplyr::mutate(chr = chr, pos = positions, mut_pos = pos, gene = gene) %>%
                dplyr::select(chr, pos, mut_pos, gene, dplyr::all_of(nucleotides))

            }) %>%
            dplyr::bind_rows() %>%
            dplyr::rowwise() %>%
            dplyr::mutate(
              total = sum(dplyr::across(dplyr::all_of(nucleotides)))) %>%
            dplyr::relocate(dplyr::all_of(nucleotides), .after = dplyr::last_col())

@@

```{r plot, fig.width = 20}
# plot positions
unique(sites$gene) %>%
  purrr::walk(function(i) {
    p <-
      sites %>%
      dplyr::filter(gene == i) %>%
      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                          values_to = 'nt_n_reads', names_to = 'nt') %>%
      dplyr::mutate(
        strandless_nt = 
          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
          forcats::fct_relevel(unique(c(paste0(toupper(nucleotides), '_ref'),
                                        paste0(toupper(nucleotides), '_alt'))))) %>%
      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
      geom_vline(aes(xintercept = mut_pos)) +
      geom_col(position = 'stack') +
      theme_bw() +
      theme(axis.text.x = element_blank()) +
      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free') +
      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
                                   C_ref = 'darkblue', C_alt = 'lightblue',
                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
                                   T_ref = 'darkred', T_alt = 'salmon',
                                   INS_alt = 'black', DEL_alt = 'grey')) +
      ggtitle(i)
    print(p)
    })
```
